---
title: "About LOINC and getting started with {roinc}"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About LOINC and getting started with {roinc}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(roinc)
```

## About LOINC

[Logical Observation Identifiers Names and Codes](https://loinc.org) or [LOINC](https://loinc.org) is a global standard providing universal codes for laboratory and clinical results. It enables interoperability between diverse health systems by ensuring data can be pooled and understood.

LOINC provides standardised, universal names and codes that support the exchange of orders and clinical observations across information systems using established syntax standards, including [HL7](https://hl7.org/fhir/), [CEN TC251](https://www.ehealth-standards.eu/centc251/), [ISO TC215](https://www.iso.org/committee/54960.html), [ASTM(https://www.astm.org/)], and [DICOM](https://www.dicomstandard.org/). These identifiers can serve as coded representations of observations within any interoperability standard that follows the observation–value model, including messaging formats, electronic documents, and application programming interfaces (APIs).

For example, LOINC codes are widely used in the Observation Identifier field (OBX-3) of the OBX segment within HL7 Observation Result (ORU) messages, including HL7 version 2.x and ASTM 1238-9410 messages. This enables standardized identification of laboratory observations when information is exchanged between systems such as Clinical Laboratory Information Management Systems (LIMS) and Electronic Health Record (EHR) systems. By providing consistent identifiers for clinical concepts, LOINC facilitates the reliable exchange and interpretation of clinical data across heterogeneous information systems.

LOINC’s database of codes can be accessed in different ways including accessing the content programmatically via application programming interfaces (APIs) to download the entire LOINC database of codes, search the entire LOINC database, and the LOINC terminology service powered by HL7 FHIR.

## The {roinc} package

The `{roinc}` package has been developed to primarily interface with these LOINC API endpoints.

The name of the package is a play between two ideas. First is that of a common approach by package developers of adding an *r* to a regular word that one is considering for a package name to make a link to this being an R package. Second is the concept of *lallation* which is a context in which someone has difficulty in enunciation such that when *l* is pronounced it sounds like an *r* or when *r* is pronounced is sounds like an *l*[^1]. So, we see the spelling of the package as *roinc* with an *r* but our recommended pronunciation is *loinc* which is the name of the standard of which it's API is what this package is interfacing with.

### What does the package do?

Currently, the package provides functions for all the LOINC API endpoints. These functions support:

* interfacing with the [LOINC download API endpoints](https://loinc.org/kb/api/download);

* interfacing with the [LOINC search API endpoints](https://loinc.org/kb/api/search-api); and,

* interfacing with the [LOINC FHIR terminology service API endpoints](https://loinc.org/fhir).

### Installation

`{roinc}` is not yet on [CRAN](https://cran.r-project.org) but can be installed from the [Oxford
iHealth R Universe](https://oxfordihtm.r-universe.dev) as follows:

```r
install.packages(
  "roinc",
  repos = c('https://oxfordihtm.r-universe.dev', 'https://cloud.r-project.org')
)
```

or via the `{pak}` package as follows:

```r
if(!require(pak)) install.packages("pak")
pak::pak("OxfordIHTM/roinc")
```

Once installed, `{roinc}` can then be loaded into an R session

```{r load-roinc}
# load package
library(roinc)
```

### Usage

#### Authentication

All requests to LOINC’s RESTful API endpoints require [basic authentication](https://en.wikipedia.org/wiki/Basic_access_authentication) using a valid LOINC username and password. LOINC logins are available by [registering on the website](https://loinc.org/join). Once you have a LOINC username and password, store these credentials as variables in your working environment by editing your `.Renviron` file and adding the following:

```bash
LOINC_USERNAME=YOUR_USERNAME
LOINC_PASSWORD=YOUR_PASSWORD
```

`{roinc}` provides a convenience function that will allow you to set these variables in your current environment as follows:

```R
roinc_set_credentials(username = "YOUR_USERNAME", password = "YOUR_PASSWORD")
```

To check whether your authentication credentials have been set, use

```R
Sys.getenv("LOINC_USERNAME")
Sys.getenv("LOINC_PASSWORD")
```

Your username and password should show up when you run these commands.

Once these environment variables are set, all functions that interface and make requests against the LOINC API endpoints will use these credentials to authenticate.

#### Retrieve LOINC metadata

Metadata on all the versions of LOINC can be retrieved via:

```{r loinc-metadata, eval = FALSE}
roinc_get_metadata()
```

which results in a `tibble` that includes information on the LOINC version, release date, number of LOINCs, and download URL sorted starting from the most current LOINC version (as shown below):

```{r loinc-metadata-show, echo = FALSE}
roinc_get_metadata()
```

If you would prefer to have the result in an unflattened list format as produced by `jsonlite::fromJSON()`, set the argument `tabular = FALSE` as follows:

```{r loinc-metadata-list, eval = FALSE}
roinc_get_metadata(tabular = FALSE)
```

which gives (showing only first 2 elements of list):

```{r loinc-metadata-list-show, echo = FALSE}
roinc_get_metadata(tabular = FALSE)[1:2]
```

If you would like to get the metadata of a specific LOINC version, for example the latest version which is 2.83, set the argument `version = "2.83"`:

```{r loinc-metadata-version, eval = FALSE}
roinc_get_metadata(version = "2.83")
```

and this will provide the following `tibble` output:

```{r loinc-metadata-version-show, echo = FALSE}
roinc_get_metadata(version = "2.83")
```

#### Download LOINC primary artifact

The LOINC primary release artifact can be downloaded. The `roinc_download()` functions allows for this download to be done programmatically as shown below:

```{r loinc-download}
roinc_download()
```

The result is the file path to the downloaded release artifact in zip format. The download is saved in the current working directory by default and the file is always named with the label `loinc_` followed by the label for the version. By default, the latest version is downloaded hence the full file path of the downloaded artifact is `loinc_latest.zip`.

We can check that the download has indeed been stored by checking the existence of the specific file in the current directory:

```{r loinc-download-check}
file.exists("loinc_latest.zip")
```

If you want to download a specific version of the LOINC primary release artifact into a different directory, you can specify the `version` and the `directory` arguments. In the following example, LOINC version 2.80 will be downloaded in a temporary directory:

```{r loinc-download-version}
roinc_download(version = "2.80", directory = tempdir())
```

We can check whether version 2.80 has indeed been downloaded in the temporary directory:

```{r loinc-download-version-check}
file.exists(file.path(tempdir(), "loinc_2.80.zip"))
```

#### Search LOINC database

You can search the extensive LOINC database using the `roinc_search_*()` set of functions. Each of these functions searches within a specific scope of the LOINC database. These scopes can either be the *LOINCs*, the *LOINC answer lists*, the *LOINC parts*, or the *LOINC groups*. For example, if you want to look up *glucose* from the LOINCs search, that can be done as follows:

```{r loinc-search, eval = FALSE}
roinc_search_loincs(query = "glucose")
```

which gives the following `tibble` of search results:

```{r loinc-search-show}
roinc_search_loincs("glucose")
```

By default, `roinc_search_loincs()` (and the other `roinc_search_*()` functions) returns a `tibble` of the first 50 rows of the search results with 47 variables/fields. These results can be customised by specifying the arguments for the *number of rows to return*, the *page of results to return*, the *variable/field to sort the results on*, and the *language for the returned results*.

#### LOINC terminology services

The LOINC database can be accessed through a terminology service defined by [HL7's FHIR](https://hl7.org/fhir/) standard. The API offers endpoints for three FHIR resources: *code system*, *value set*, and *concept map*. It also has a set of endpoints for *questionnaire* resources to LOINC's wealth of standard assessments and survey content including all child concepts (i.e. questions) and answer lists of a given assessment.

##### Code system

To get properties available for LOINC as a CodeSystem, use the following command:

```{r loinc-codesysem, eval = FALSE}
roinc_codesystem()
```

which returns:

```{r loinc-codesystem-show, echo = FALSE}
roinc_codesystem()
```

To get information on an individual LOINC term, Part, Answer List, or Answer string, use `roinc_codesystem_lookup()`. For example, using the code `4544-3` for hematocrit:

```{r loinc-codesystem-lookup, eval = FALSE}
roinc_codesystem_lookup(code = "4544-3")
```

which returns an unflattened list format as produced by `jsonlite::fromJSON()`.

To get a specific piece of information, specify the argument for `property` using any of the different variables/fields available from the system. For example, for the code `4544-3` for hematocrit and retrieving information on which version of LOINC this was first released:

```{r loinc-codesystem-lookup-property}
roinc_codesystem_lookup(code = "4544-3", property = "VersionFirstReleased")
```

##### Value sets

FHIR’s *ValueSet* resource specifies a set of codes drawn from one or more code systems. In LOINC’s case, three main types of value sets are defined:

* Broadly useful collections of LOINC terms
* Answer Lists, e.g. LL4700-2
* LOINC Groups, e.g. LG32763-1

To return a value set definition, `roinc_valueset()` can be used. For example, to get the value set definition for answer list *LL1162-8*:

```{r loinc-valueset, eval = FALSE}
roinc_valueset(valueset = "LL1162-8")
```

which returns

```{r loinc-valueset-show, echo = FALSE}
roinc_valueset(valueset = "LL1162-8")
```

To get the list of terms in a value set, specify the argument `expand = TRUE` as follows:

```{r loinc-valueset-expand, eval = FALSE}
roinc_valueset(valueset = "LL1162-8", expand = TRUE)
```

which returns

```{r loinc-valueset-expand-show, echo = FALSE}
roinc_valueset(valueset = "LL1162-8", expand = TRUE)
```

To determine if a term is in a given ValueSet, `roinc_valueset_validate()` can be used. For example, to determine if the LOINC term `8867-4` is in the LOINC group `LG33055-1`:

```{r loinc-valueset-validate, eval = FALSE}
roinc_valueset_validate(valueset = "LG33055-1", code = "8867-4")
```

which returns:

```{r loinc-valueset-validate-show, echo = FALSE}
roinc_valueset_validate(valueset = "LG33055-1", code = "8867-4")
```

In the result, the `valueBoolean` returns a value `TRUE` which means that the term is included in the specified value set.



[^1]: A condition that the author of the package suffered from at their young formative years.